\textit{MSF: Modulated Sub-graph Finder}

### 2019-03-08 MSF_Paper_FA.tex
jActiveModules improved

	\DONE{Recently systems biology approaches are also helping illuminate the structural and functional design of a cell~\cite{mitra_carvunis_ramesh_ideker_2013}. Mitra and colleagues discuss the different systems biology approaches to identify and measure the modules of inter-molecular interactions steering a particular biological process. These approaches are classified into four categories, identification of "active modules" with molecular profiles, identification of "conserved modules" across species , identification of "differential modules" with multiple conditions and identification of "composite modules" with interactions of different types. Identification of the active modules is one of the most successful approaches, which overlays the modules or networks onto the molecular profiles (transcriptomics, genomics, proteomics) to identify the "active modules". To find these active modules heuristics solutions like simulated annealing (SA), greedy methods, genetic algorithms (GA), network propagation and co-clustering methods are used~\cite{mitra_carvunis_ramesh_ideker_2013}. JActiveModules~\cite{jActiveModules} has been the first of this kind using simulated annealing to find the active modules.}

	\DONE{The motivation was to find the modulated biological processes cross-talking, for which identifying the modulated networks first with the help of systems biology was necessary.} On these grounds we propose a novel approach to make use of the rich gene
	and protein interaction annotation resources available, to gain additional
	functional insights from basic DGEA. To this, we start with the
	presupposition that expression of neighboring genes within a functional
	pathway are not independent from each other. Rather, they are often
	regulating each others expression or are part of the same
	regulon~\cite{Michalak}. We understand that the categorization of links
	between genes into labeled pathways is often an arbitrary one, given the
	extensive cross talk between different pathways. Although these categories
	have proven to be useful in many situations, they force a certain
	perspective onto the interpretation of novel data. Based on these two
	principles, we aim to find sub-graphs of connected genes within cell
	signaling network, which exhibit as a whole significant differential
	expression changes. This approach differs in two main aspecarg1ts from common
	pathway analysis tools. First, it does not aim to identify functional
	pathways enriched in differentially expressed genes, but detects sub-graphs
	or branches in a network graph (potentially spanning more than one
	functionally grouped pathway) which is coherently modulated. Second, it
	aims to improve the DGEA on the gene level, by collecting the information
	of neighboring genes, which as a whole might exhibit prominent enough
	signal to be called; again as a whole, significantly modulated. All of this can be helpful to understand the cause and effect of a stimulus and might inform about potential points of intervention.

	As input, information on functional links between genes provided by
	e.g.~KEGG or Reactome and information on the differential expression status
	of single genes resulting from a DGEA, are required. As a result the
	analysis returns sub-graphs and their joint confidence scores, reflecting
	how the perturbation is migrated through the network. Furtharg1ermore, the
	entry points of perturbation in the networks and overlap with conventional
	pathway categories are returned. These entry points are given impact scores showing the degree of perturbations introduced by them. The reliability of the entry points is checked using a T-test by investigating the significant difference between the downstream and upstream genes of each entry point. The output is prepared in a directed
	adjacency file, convenient for visualization, e.g., with
	StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

	The proposed algorithm is named
	\texttt{Modulated Sub-graph Finder} (abbreviated
	\texttt{MSF}). \texttt{MSF} can help transform the information obtained
	from DGEA into comprehensible knowledge of signal transduction of genes and
	thereby being a valuable complement to existing pathway based
	methods.  \texttt{MSF} is freely accessible on github under the terms of the
	Creative Commons Attribution 4.0 International License.

@@

	Lambda $\lambda$ be the weights for each gene, currently all genes have equal weight i.e. 1. The combined \textit{p}-value  $t(\rho)$ of a sub-graph will express the
	significance of all genes in the sub-graph being modulated
	together. Thereby, the information from the different genes are used as,
	although not independent, replicated measurement of the behavior of the
	whole sub-graph. This potentially increases the power to detect also
	significant sub-graphs consisting of genes which are not significant on
	there own.
	\newline

@@

	In the next step, we check if any of the initial modulated
        sub-graphs could further be extended beyond the immediate
          neighbouring genes. Instead of testing single genes and
          their compatibility to be added, groups of genes are
          considered. If the combined \textit{p}-value of the
        initial modulated sub-graph and the extension genes is smaller
        than the \textit{p}-value of the initial sub-graph the
        extension is accepted. All possible extension paths up to
        \emph{3} (default 2) genes at all nodes in the sub-graph are
        tested. Again, this step is iteratively repeated until no
        further genes are added to the significant differentially
        expressed sub-graphs. This step bridges small gaps of genes
        without a clear differential signal in the DGEA.\newline

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, expressing the relative number of downstream genes
          within the corresponding sub-graph directly connected by
          directed links. This score can be interpreted as an upper
          limit of how much of the sub-graph's perturbation could have
          been introduced by the respective source, and thereby could
          be helpful to prioritize different identified sources for
          larger sub-graphs. In the same
        spirit as sources, the most downstream genes of the
        modulated sub-graphs are identified and defined as sinks. Due
        to loops not all sub-graphs are guaranteed to have sources or
        sinks. The reliability of each source is inspected using a T-test. The significant difference between the two groups, genes downstream the source to the genes upstream of the source is determined. This would help to assess if the source identified is reliable and marks the border between two different regulation regimes.

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. Additionally, a
          file containing details on all sources and sinks for all
          modulated sub-graphs is reported. Furthermore, for
          facilitated visualisation in Cytoscape, a node attribute
          file is provided, containing the source weightage and the
          log-fold chances of all considered genes.

@@

	\DONE{The only system requirements to run MSF are Java version 8 and JDK 1.8. The few package
		dependencies are already been added to the release. To run
		MSF, the user must provide two text files, one containing the
		DGEA from edgeR~\cite{edgeR} or DESeq2~\cite{love2014moderated}} and  the second one containing
        directed interactions in an adjacency format file. Example
        files and a detailed tutorial to use MSF has been provided on
        github https://github. com/Modulated-Subgraph-Finder/MSF.

@@

	To demonstrate its usefulness, \texttt{MSF} is applied to
        RNA-seq data set of primary human monocyte-derived macrophages
        (MDMs) infected with Ebola virus
        (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
        Filoviridea family: filamentous, enveloped and single stranded
        RNA viruses. EBOV causes hemorrhagic fever in humans, inducing
        the host innate and adaptive immune response to be unable to
        control virus infection~\cite{Prins}. Currently, there are no
        approved antiviral drugs for the treatment of Ebola virus
        infection~\cite{Konde,Rhein}.  The initial targets of EBOV are
        the macrophages and dendritic immune
        cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate
        immune response of the host, which includes the activation of
        alpha/beta interferon (IFN-$\alpha /
        \beta$)~\cite{Prins,Konde,Cardenas}. It has been proposed that
        IFN-$\alpha / \beta$ should be tested against Ebola for its
        antiviral activity through clinical trials~\cite{Konde}. Ebola
        infection data was selected to test the approach because it
        has been well recognized for the last several decades, and
        vast literature is available for the pathogenesis of Ebola,
        thereby, facilitating the verification of the results of
        \texttt{MSF} with the vast literature present on Ebola
        infection. Especially, the detection of IFN-$\alpha / \beta$
        as point of action for the virus, could be considered as a
        basic indicator of the correctness and usefulness of the
        approach.

@@

	For the earliest time point at 6~hpi, three large modulated
        sub-graphs were identified with 42, 139, and 69 genes.  The
        modulated sub-graphs consist predominantly of cytokines,
        chemokines (CXCL10, CCL8, CXCL9, CXCL11, CXCR4, CCR7, CCL4L1,
        CCL3L1, CCL4, CCL8, CCL20, CCL3, CCL19) and Interleukin genes
        (IL6, IL27, IL23).  IFNB1 and IFNA1 were both identified as
        two of the possible sources in the most significantly
        modulated sub-graph identified with 42 genes. The impact score
        of IFNB1 is 14.5\%, \DONE{the highest impact score and IFNA1 is 8.7\%, in top 5 highest impact scores in the sub-graph they
        	belong to} . Most of the sources identified by \texttt{MSF}
        were type~I interferon induced genes (Supplement material).
        At 24~hpi seven modulated sub-graphs were identified with four
        main sub-graphs consisting of 61, 222, 130 and 242 genes,
        others being smaller than 6 genes. Again, IFNB1 and IFNA1 were
        identified as two sources out of the total sources with 3.9\% 
        and 1.6\% impact score. \DONE{IFNB1 was one of the top 5 impact score sources for the corresponding sub-graph}. For the last time-point 48~hpi, six
        modulated sub-graphs were identified. Three of the sub-graphs
        were less than ten genes and main sub-graphs had 217, 224 and
        276 genes. IFNB1 and IFNA1 were identified as sources in the
        most significantly modulated sub-graph with an impact score of
        2.8\% and 3.7\%, \DONE{but not the highest ranked sources} (Supplement material).

@@

	Three main modulated sub-graphs identified by \texttt{MSF} at
        6~hpi are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graphs
        represent the immediate output of the \texttt{MSF}-analysis,
        visualized by StringApp~\cite{StringApp} in
        Cytoscape~\cite{Cyto}. Each node represents a gene part of a
        modulated sub-graph, whereby the associated colors code the
        functional annotation deduced from KEGG Pathways. The
        cross-talk between the pathways and also the multiplhttps://github.com/Modulated-Subgraph-Finder/MSF/tree/master/bine
        employment of many genes is evident. The flow of information
        between the sensors and effectors can be perceived given the
        directionality of each interaction, indicated by arrows.  In
        more detail, sub-graph~1 (bottom) shows how the activation of
        toll-like receptor, cytokine, chemokine activating jak-stat
        and mapk genes, together with TNF leads into apoptosis. The
        next significant sub-graph (sub-graph~2: top right) reveals
        how information from the Extra-cellular matrix (ECM) receptor,
        which are reported to interact with Ebola glycoprotein
        (GP)~\cite{Veljkovic}, chemokines, cytokines, and cytosolic
        DNA sensing are directly or indirectly controlling cell
        growth, differentiation, proliferation and apoptosis. It
        suggests that dysregulation of these pathways is responsible
        for modulation of apoptosis. Eventually, sub-graph~3 (top
        left) demonstrates how INFA1 and INFB1 modulates onchttps://github.com/Modulated-Subgraph-Finder/MSF/tree/master/bine more,
        via only a few intermediate steps, the apoptotic response of
        the cell. On the other side cAMP signaling genes activates
        platelet genes.

@@

	\subsection*{Robustness}

	A potential concern is how noise in the gene expression
        measurements affects our analysis. To assess the robustness
        and stability of our method, we therefore added Poisson
        distributed noise to the read counts of the three time-points
        data set, used above.  Then DGEA was carried out on the
        disturbed data with the same parameters as for the native data
        using edgeR, followed by analysis with \texttt{MSF}. This
        procedure was carried out 100 times.  Every time the genes
        from the modulated sub-graphs identified from noisy data were
        compared to the genes of sub-graphs identified from the native
        data. The robustness of \texttt{MSF} analysis for the
        time-point 6, 24, and 48~hpi is shown in
        Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was
        modeled can be considered as rather stringent as \texttt{MSF}
        is sensitive to \textit{p}-value changes across the whole
        range of possible values. The observed median recall rates lay
        between 71~\% (6~hpi) and 84~\% (48~hpi).

@@

	 \DONE{The purpose of benchmark is mainly to show overall capabilities of \texttt{MSF}. \texttt{MSF} was compared to jActivemodules~\cite{jActiveModules} since they use similar approach to find and score the modules. Comparison of \texttt{MSF} to Reactome~\cite{Reactome} was done because  Reactome is the standard tool used when looking for modulated pathways after DGEA. Since \texttt{MSF} and GSEA~\cite{Subramanian15545} use no cut-off to the DGEA, their results were also compared.}

@@

	jActiveModules~\cite{jActiveModules} is a plugin in Cytoscape that searches for molecular
        interaction network to find expression activated
        sub-networks. The method used to score the expression
        activated sub-networks is close to the method used in
        \texttt{MSF}.  \DONE{The difference is in how these sub-networks are identified. \texttt{MSF} starts building the sub-graphs from one gene, incorporating and combining the \textit{p}-value of the next gene, with the check that the combined \textit{p}-value of new sub-graph should be better than the original. On the other hand jActiveModules first transforms all the gene’s \textit{p}-values to z-scores and tries to find connected sets of genes with unexpectedly high levels of differential expression, in this case high z-scores. The overall score of the sub-network is calculated by combining the z-scores of the genes. Next using their extended simulated annealing method jActiveModules finds the user defined number of highest scoring putative modules by toggling multiple nodes to merge additional components.}

	The first time-point of Ebola infection data was analyzed
        using jActiveModules (Version 3.2.1) to compare the modulated
        sub-graphs identified by \texttt{MSF} and jActiveModules. The
        input files were same for both tools, number of modules
        parameter was set to 5 (default). This would report 5 putative modules even when only one of them is good. Out of the 5 putative modules identified by jActiveModules, the module with the highest pathScore was selected for comparison with
        \texttt{MSF} identified modulated sub-graphs. The module consists of a single graph with 314
        genes. While \texttt{MSF} identified three directed modulated
        sub-graphs with 42, 69 and 139 genes. The overlap of the
        common genes identified between \texttt{MSF} and jActiveModules
        is shown in Fig~\ref{fig:MSFvsJActiveModule}. \DONE{JActiveModules identified a single module, since this module had unexpectedly high levels of differential expression. With the extended simulated annealing method, jActiveModules toggled multiple nodes to merge a number of components with low score into this single high-scoring component making it larger than original.} Although jActiveModules was able to identify
        single connected module, \texttt{MSF} provides directionality,
        with the identification of possible perturbation sources of
        the sub-graphs.

@@

	All enriched pathways with a cut-off of \textit{p}-value
        < 0.05 for \texttt{MSF} and DEG list for the
        three time-points were selected. The comparison shows most of
        the pathways known from literature to be dis-regulated by
        Ebola infection are enriched in both the enrichment
        analysis. Toll-Like receptor signaling pathway when interacts
        with EBOV glycoprotein (GP), it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

	\textbf{Gene set enrichment analysis (GSEA)}

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll\FA{-}like recpetor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,arg1
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how
        the cross-talk and topology of genes between the pathways
        identifed. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\subsection*{Supplement material}
	Supplementary material is available form GitHub: https://github.com/Modulated-Subgraph-Finder/MSF

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{17 cm}
				\includegraphics[width=1.0\linewidth]{AlgorithmMSF.png}
				\caption{Graphical representation of the \texttt{MSF}
					heuristical approach to detect modulated sub-graphs
					in a global gene regulatory network.}
				\label{fig:pseudocodemsf}
			\end{minipage}
		}

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{18 cm}
				\includegraphics[width=1.0\linewidth]{DirectedMerged_Network.png}
				\caption{Visualisation of the three modulated
					directed sub-graphs identified by \texttt{MSF} at 6~hpi. The
					node coloring is associated to KEGG pathways
					referring to the colors in the legend. The
					graph edges are from Reactome. \label{fig:Sub-graph6hpi}}
			\end{minipage}
		}
	\end{figure*}

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{12 cm}
				\includegraphics[width=0.9\linewidth]{MSFvsJActiveModule.png}
				\caption{The Venn diagram shows the common genes identified as modulated from \texttt{MSF} identified sub-graphs and jActiveModule identified module.}
				\label{fig:MSFvsJActiveModule}
			\end{minipage}
		}
	\end{figure*}

>>>>

	From another perspective, network based approaches do not categorize sets of genes into functional pathways, but they consider all interactions to be equal. Thereby, they avoid distinguishing arbitrarily between interactions within a pathway and interactions between pathways (i.e., cross-talk). With this they aim to identify sub-networks that show modulation between two conditions or upon a stimuli~\cite{mitra_carvunis_ramesh_ideker_2013}. To find these active modules heuristics solutions like simulated annealing (SA), greedy methods, genetic algorithms (GA), network propagation and co-clustering methods are used~\cite{mitra_carvunis_ramesh_ideker_2013}. jActiveModules has been the first of this kind using simulated annealing to find the modulated sub-networks~\cite{jActiveModules}. The benefit of omitting pathways is bought by reduced interpretability of the results due to the lack of functional labels on the networks.

    On these grounds we propose a novel approach to make use of the rich gene
	and protein interaction annotation resources available and combining it to functional pathway annotations to gain additional insights from basic DGEA. To this, we start with the
	presupposition that expression of neighboring genes within a functional
	pathway are not independent from each other. Rather, they are often
	regulating each others expression or are part of the same
	regulon~\cite{Michalak}. We understand that the categorization of links
	between genes into labeled pathways is often an arbitrary one, given the
	extensive cross talk between different pathways. Although these categories
	have proven to be useful in many situations, they force a certain
	perspective onto the interpretation of novel data. Based on these two
	principles, we aim to find sub-graphs of connected genes within cell
	signaling network, which exhibit as a whole significant differential
	expression changes. This approach differs in two main aspects from common
	pathway analysis tools. First, it does not aim to identify functional
	pathways enriched in differentially expressed genes, but detects sub-graphs
	or branches in a network graph (potentially spanning more than one
	functionally grouped pathway) which is coherently modulated. Second, it
	aims to improve the DGEA on the gene level, by collecting the information
	of neighboring genes, which as a whole might exhibit prominent enough
	signal to be called; again as a whole, significantly modulated. All of this can be helpful to understand the cause and effect of a stimulus and might inform about potential points of intervention.

	As input, information on functional links between genes provided by
	e.g.~KEGG or Reactome and information on the differential expression status
	of single genes resulting from a DGEA, are required. As a result the
	analysis returns sub-graphs and their joint confidence scores, reflecting
	how the perturbation is migrated through the network. Furthermore, the
	entry points of perturbation in the networks and overlap with conventional
	pathway categories are returned. These entry points are given impact scores showing the degree of perturbations introduced by them. The reliability of the entry points is checked using a t-test on the \textit{p}-values by investigating the significant difference between the downstream and upstream genes of each entry point. The output is prepared in a directed
	adjacency file, convenient for visualization, e.g., with
	StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

	The proposed algorithm is named
	\texttt{Modulated Sub-graph Finder} (abbreviated
	\texttt{MSF}). \texttt{MSF} can help transform the information obtained
	from DGEA into comprehensible knowledge of signal transduction of genes and
	thereby being a valuable complement to existing pathway based
	methods. \texttt{MSF} is freely accessible on git hub under the terms of the
	Creative Commons Attribution 4.0 International License.

@@

	Lambda $\lambda$ be the weights for each gene, currently all genes have equal weight i.e. 1. The combined \textit{p}-value  $t(\rho)$ of a sub-graph will express the
	significance of all genes in the sub-graph being modulated
	together. Thereby, the information from the different genes are used as,
	although not independent, replicated measurement of the behavior of the
	whole sub-graph. This potentially increases the power to detect also
	significant sub-graphs consisting of genes which are not significant on
	there own.

@@

	In the next step, we check if any of the initial modulated
        sub-graphs could further be extended beyond the immediate
          neighboring genes. Instead of testing single genes and
          their compatibility to be added, groups of genes are
          considered. If the combined \textit{p}-value of the
        initial modulated sub-graph and the extension genes is smaller
        than the \textit{p}-value of the initial sub-graph the
        extension is accepted. All possible extension paths up to
        \emph{3} (default 2) genes at all nodes in the sub-graph are
        tested. Again, this step is iteratively repeated until no
        further genes are added to the significant differentially
        expressed sub-graphs. This step bridges small gaps of genes
        without a clear differential signal in the DGEA.\newline

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, expressing the relative number of downstream genes
          within the corresponding sub-graph directly connected by
          directed links. This score can be interpreted as an upper
          limit of how much of the sub-graph's perturbation could have
          been introduced by the respective source, and thereby could
          be helpful to prioritize different identified sources for
          larger sub-graphs. In the same
        spirit as sources, the most downstream genes of the
        modulated sub-graphs are identified and defined as sinks. Due
        to loops not all sub-graphs are guaranteed to have sources or
        sinks. The reliability of each source is inspected using a t-test on the \textit{p}-values . The significant difference between the two groups, genes downstream the source to the genes upstream of the source is determined. This would help to assess if the source identified is reliable and marks the border between two different regulation regimes.

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. Additionally, a
          file containing details on all sources and sinks for all
          modulated sub-graphs is reported. Furthermore, for
          facilitated visualization in Cytoscape, a node attribute
          file is provided, containing the source weight-age and the
          log-fold chances of all considered genes.

@@

	The only system requirements to run \texttt{MSF} are Java version 8 and JDK 1.8. The few package
		dependencies are already been added to the release. To run
		\texttt{MSF}, the user must provide two text files, one containing the
		DGEA and the second one containing
        directed interactions in an adjacency format file. Example
        files and a detailed tutorial to use MSF has been provided on
        git hub \url{https://github.com/Modulated-Subgraph-Finder/MSF}.

@@

	To demonstrate its usefulness, \texttt{MSF} is applied to
        RNA-seq data set of primary human monocyte derived macrophages
        (MDMs) infected with Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
        Filoviridea family: filamentous, enveloped and single stranded
        RNA viruses. EBOV causes hemorrhagic fever in humans, inducing
        the host innate and adaptive immune response to be unable to
        control virus infection~\cite{Prins}. Currently, there are no
        approved antiviral drugs for the treatment of Ebola virus
        infection~\cite{Konde,Rhein}.  The initial targets of EBOV are
        the macrophages and dendritic immune
        cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate
        immune response of the host, which includes the activation of
        alpha/beta interferon (IFN-$\alpha /
        \beta$)~\cite{Prins,Konde,Cardenas}. It has been proposed that
        IFN-$\alpha / \beta$ should be tested against Ebola for its
        antiviral activity through clinical trials~\cite{Konde}. Ebola
        infection data was selected to test the approach because it
        has been well recognized for the last several decades, and
        vast literature is available for the pathogenesis of Ebola,
        thereby, facilitating the verification of the results of
        \texttt{MSF} with the vast literature present on Ebola
        infection. Especially, the detection of IFN-$\alpha / \beta$
        as point of action for the virus, could be considered as a
        basic indicator of the correctness and usefulness of the
        approach.

@@

	For the earliest time point at 6~hpi, three large modulated
        sub-graphs were identified with 42, 139, and 69 genes.  The
        modulated sub-graphs consist predominantly of Cytokines,
        chemokines (CXCL10, CCL8, CXCL9, CXCL11, CXCR4, CCR7, CCL4L1,
        CCL3L1, CCL4, CCL8, CCL20, CCL3, CCL19) and Interleukin genes
        (IL6, IL27, IL23).  IFNB1 and IFNA1 were both identified as
        two of the possible sources in the most significantly
        modulated sub-graph identified with 42 genes. The impact score
        of IFNB1 is 14.5, the highest impact score and IFNA1 is 8.7, in top 5 highest impact scores in the sub-graph they belong to. Most of the sources identified by \texttt{MSF}
        were type~I interferon induced genes (Supplement material).
        At 24~hpi seven modulated sub-graphs were identified with four
        main sub-graphs consisting of 61, 222, 130 and 242 genes,
        others being smaller than 6 genes. Again, IFNB1 and IFNA1 were
        identified as two sources out of the total sources with 3.9 
        and 1.6 impact score. IFNB1 was one of the top 5 impact score sources for the corresponding sub-graph. For the last time-point 48~hpi, six
        modulated sub-graphs were identified. Three of the sub-graphs
        were less than ten genes and main sub-graphs had 217, 224 and
        276 genes. IFNB1 and IFNA1 were identified as sources in the
        most significantly modulated sub-graph with an impact score of
        2.8 and 3.7, but not among the highest ranked sources (Supplement material).

@@

	Three main modulated sub-graphs identified by \texttt{MSF} at
        6~hpi are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graphs
        represent the immediate output of the \texttt{MSF}-analysis,
        visualized by StringApp~\cite{StringApp} in
        Cytoscape~\cite{Cyto}. Each node represents a gene part of a
        modulated sub-graph, whereby the associated colors code the
        functional annotation deduced from KEGG Pathways. The
        cross-talk between the pathways and also the multiple
        employment of many genes is evident. The flow of information
        between the sensors and effectors can be perceived given the
        directionality of each interaction, indicated by arrows.  In
        more detail, sub-graph~1 (bottom) shows how the activation of
        Toll-like receptor, Cytokine, Chemokine activating Jak-STAT
        and MAPK genes, together with TNF leads into apoptosis. The
        next significant sub-graph (sub-graph~2: top right) reveals
        how information from the Extra-cellular matrix (ECM) receptor,
        which are reported to interact with Ebola glycoprotein
        (GP)~\cite{Veljkovic}, Chemokines, Cytokines, and Cytosolic
        DNA sensing are directly or indirectly controlling cell
        growth, differentiation, proliferation and apoptosis. It
        suggests that dysregulation of these pathways is responsible
        for modulation of apoptosis. Eventually, sub-graph~3 (top
        left) demonstrates how INFA1 and INFB1 modulates once more,
        via only a few intermediate steps, the apoptotic response of
        the cell. On the other side cAMP signaling genes activates
        platelet genes.

@@

	\subsection*{Robustness}Gene set enrichment analysis

	A potential concern is how noise in the gene expression
        measurements affects our analysis. To assess the robustness
        and stability of our method, we therefore added Poisson
        distributed noise to the read counts of the three time-points
        data set, used above. Then DGEA was carried out on the
        disturbed data with the same parameters as for the native data
        using edgeR, followed by analysis with \texttt{MSF}. This
        procedure was carried out 100 times.  Every time the genes
        from the modulated sub-graphs identified from noisy data were
        compared to the genes of sub-graphs identified from the native
        data. The robustness of \texttt{MSF} analysis for the
        time-point 6, 24, and 48~hpi is shown in
        Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was
        modeled can be considered as rather stringent as \texttt{MSF}
        is sensitive to \textit{p}-value changes across the whole
        range of possible values. The observed median recall rates lay
        between 71~\% (6~hpi) and 84~\% (48~hpi).Gene set enrichment analysis

@@

	 The purpose of the comparisons to existing tools is to show the overall capabilities of \texttt{MSF}. \texttt{MSF} was compared to jActiveModules~\cite{jActiveModules} since they use similar approaches to find and score modules. For comparison to classical pathway enrichment analysis Reactome~\cite{Reactome} and gene set enrichment analysis (GSEA)~\cite{Subramanian15545} was chosen since both are widely used and the latter does not rely on \textit{p}-value cut-off.

@@

	jActiveModules~\cite{jActiveModules} is a plugin in Cytoscape that searches for molecular
        interaction network to find expression activated
        sub-networks. The method used to score the expression
        activated sub-networks is close to the method used in
        \texttt{MSF}. The difference is in how these sub-networks are identified. \texttt{MSF} starts building the sub-graphs from one gene, incorporating and combining the \textit{p}-value of the next gene, with the check that the combined \textit{p}-value of new sub-graph should be better than the original. On the other hand jActiveModules first transforms all the gene’s \textit{p}-values to z-scores and tries to find connected sets of genes with unexpectedly high levels of differential expression, in this case high z-scores. The overall score of the sub-network is calculated by combining the z-scores of the genes. Next using their extended simulated annealing method jActiveModules toggles multiple nodes to merge additional components.

	The first time-point of Ebola infection data was analyzed
        using jActiveModules (Version 3.2.1) to compare the modulated
        sub-graphs identified by \texttt{MSF} and jActiveModules. The
        input files were same for both tools. From the modules identified by jActiveModules, the module with the highest pathScore was selected for comparison with
        \texttt{MSF} identified modulated sub-graphs. The module consists of a single graph with 314
        genes. While \texttt{MSF} identified three directed modulated
        sub-graphs with 42, 69 and 139 genes. The overlap of the
        common genes identified between \texttt{MSF} and jActiveModules
        is shown in Fig~\ref{fig:MSFvsJActiveModule}. The sub-graphs identified by \texttt{MSF} are more fragmented than jActiveModules. Unfortunately there is no golden standard example data that could help benchmark the method. \texttt{MSF} provides directionality,
        with the identification of possible perturbation sources of
        the sub-graphs. The predefined pathway labels could also be seen in \texttt{MSF} identified sub-graph with little effort using StringApp~\cite{StringApp}.

@@

	All enriched pathways with a cut-off of \textit{p}-value
        < 0.05 for \texttt{MSF} and DEG list for the
        three time-points were selected. The comparison shows most of
        the pathways known from literature to be dis-regulated by
        Ebola infection are enriched in both the enrichment
        analysis. Toll-Like receptor signaling pathway when interacts
        with EBOV glycoprotein (GP), it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine Toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

	\textbf{Gene set enrichment analysis }

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll-like receptor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how
        the cross-talk and topology of genes between the pathways
        identified. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\subsection*{Supplement material}
	Supplementary material is available form Git Hub: https://github.com/Modulated-Subgraph-Finder/MSF

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{17 cm}
				\includegraphics[width=1.0\linewidth]{AlgorithmMSF.png}
				\caption{Graphical representation of the \texttt{MSF}
					heuristic approach to detect modulated sub-graphs
					in a global gene regulatory network.}
				\label{fig:pseudocodemsf}
			\end{minipage}
		}

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{18 cm}
				\includegraphics[width=1.0\linewidth]{DirectedMerged_Network.png}
				\caption{Visualization of the three modulated
					directed sub-graphs identified by \texttt{MSF} at 6~hpi. The
					node coloring is associated to KEGG pathways
					referring to the colors in the legend. The
					graph edges are from Reactome. \label{fig:Sub-graph6hpi}}
			\end{minipage}
		}
	\end{figure*}

@@

	\begin{figure*}
		\centering
		\fbox{
			\begin{minipage}{12 cm}
				\includegraphics[width=0.9\linewidth]{MSFvsJActiveModule.png}
				\caption{The Venn diagram shows the common genes identified as modulated from \texttt{MSF} identified sub-graphs and jActiveModules identified module.}
				\label{fig:MSFvsJActiveModule}
			\end{minipage}
		}
	\end{figure*}

	\begin{figure*}[p]
		\centering
		\fbox{
			\begin{minipage}{15 cm}
				\includegraphics[width=1.01\linewidth]{ReactomeVsMSF.png}
				\caption{The Upset plot shows the
                                  number of shared pathways between
                                  \texttt{MSF} identified sub-graph
                                  gene list and DEG cut-off list for
                                  the three time-points. All
                                  10 different
                                  Toll-like receptor cascades are
                                   in the set
                                    of 164 shared
                                  pathways only between \texttt{MSF}
                                  at different time-points.}
				\label{fig:msfvsreactome}
			\end{minipage}
		}
	\end{figure*}
